A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-04-19 text/manuscript.Rmd
figure references added; header amended add/format TOC and style plain html

---
title: "TTmanuscript"
author: "Dennis Hazelett"
date: "4/18/2020"
output: html_document
bibliography: references.bib
---

@@

# Title: A molecular taxonomy of tumors independent of tissue-of-origin

#### Authors: Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett

## Abstract: 
Cancer is a disease that affects metabolism and other basic cellular processes common to all cells. In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying theories elude the field. To address this challenge we reduced omics mutation and expression profiling data to pathway memberships. We found that tumors can be classified according to pathway disruption in twelve distinct classes orthogonal to a tissue of origin (including histological subtypes), number of mutations and clinical stage. Moreover, known germline risk mutations such as BRCA1/2 are equally distributed among the classes. One class is distinguished by mutations in extracellular receptors and their ligands, particularly RTKs including the ERBB2/4, EGFR, Insulin Receptor and IGFR. This class is neatly subdivided into PI3K/Akt and Ras branches. Of these subclasses, the PI3K/Akt subgroup also has multiple lesions in immune-related pathways, including both innate immunity and T-cell receptor signaling. Another class is typified by mutations in the cellular response to hypoxia and defects in beta-catenin and hedgehog signaling. Intriguingly, we identify a novel class of mutations in proteasome component proteins that may be sufficient to cause compound defects across multiple signaling pathways in this class. A third class is typified by defects in metabolism, especially of RNA and protein biosynthesis. In addition to pathways that distinguish between the 3 major classes of tumor, we identify common mutations across tumors and stages, as well as pathways/mutations that distinguish between metastatic tumors and non-metastatic tumors. With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found. We discuss the implications for tumor biology and treatment.

## Introduction:

@@

## Results:

### Classification of tumors by pathway disruption reveals 3 major groups

@@

We noted intriguing similarities across different clusters, which motivated us to use dimensional reduction with UMAP (arXiv:1802.03426). Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings (Figure 1A). We call these tentatively by roman numerals I-V. While many of the clusters from the prior analysis remained close in proximity (note the colors in Figure 1A), some redistribution occurred, potentially reflecting important differences that could not be captured using the hierarchical method. Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14). To determine whether the overall structure of the graph was driven by mutation counts, we examined the distribution of mutations (for pathway genes only). While there is a subtle trend, mostly visible for groups I and V (violin plot, Figure 1B), mutation count does not appear to be the dominant feature. To test whether the 5 groups represent progression by stage (which are determined clinically by size, though multiple inconsistent staging paradigms exist), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every grouping about equally (Figure 1C).

The 15 hierarchical clusters were largely free of any tissue-of-origin bias, so we suspected that our 5 groups would be similarly independent from tissue of origin. To visualize tissue of origin relative to the groups we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the UMAP, and did not observe any bias to group by these tissues of origin (Figures 1D). Full projections for all 23 cancers are presented in Supplementary Figure S2. Some cancers are defined by multiple subtypes of molecular or histologically distinguishing features, which are often the basis for different treatment therapies. One hypothesis to explain the lack of bias by cancer type is that the groupings are representative of these molecular subtypes. To probe this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP. Perhaps surprisingly we did not detect any group biases for these or any other subtype annotations in GDC (see Supplementary Figure S3 for the full set of projections), suggesting that these subtypes are reflective of differences either in specialized cell-type-of-origin or other cancer risk mechanisms.

There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes. However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length. BRCA1/2 genes for example are key for DNA double-stranded break repair (ref) and confer risk for breast, prostate and ovarian cancers (refs). Since the mechanism of risk is thought to involve loss of heterozygosity, we projected the somatic mutations for BRCA1 and 2 genes onto the UMAP, but did not observe any segregation of these mutations by group (Figure 1E). We made similar projections for the MMR genes MSH2, MSH6, MLH, PMS1 and PMS2; BRIP1, RAD15, CHEK2 and APC. None of these genes exhibited any remarkable specificity with respect to group (except maybe APC??) (Supplementary Figure S4). To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the TET2 and TET3 genes, plus telomerase 1, 2, 3, and observed similar lack of segregation by group (Supplementary Figure S4).

### Determinants of Tumor Class

In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 381 pathways used to produce it (Supplementary Figure S5), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment. We visualized the enrichment as a heatmap, showing only the top significant pathways by x-fold enrichment (Figure 2A). Using this approach, we identified 6? pathways associated with class II, 34 pathways primarily enriched in group III tumors, and 10 pathways enriched in class IV tumors. Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively. Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (Figure 2B). Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution. Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (Figure 2C). Class IV had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, beta-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others. As groups III and IV were the most distinct of the 5, we chose to investigate these two groups further.

Group III contained interesting sub-patterns of enrichment, which we subdivided further into class IIIa and class IIIb sequentially from right following the global naming system. Common to classes IIIa and b were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as EGFR, IGF1R, Insulin receptor, and ERBB2 and 4. In addition to these we observed SCF-Kit, DAP12, and Tie 2 signaling pathways, altogether suggesting that Class III tumors share mutations in extracellular signaling pathways generally. Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment. We plotted mutations at the gene level, grouped in a supervised manner by pathway membership, and then performed unsupervised clustering of the tumors (Figure 3 or supplement figure?). We find….(please do this analysis)

@@

### Shared mutated genes/pathways

### Oncoprint stuff

### Determinants of Metastatic vs non-metastatic tumors

Following the same logic we used to investigate cluster-specific enrichment of pathways (Figure 2) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (Figure 1C). Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown). We reasoned that the clusters might be different enough to dilute cluster-specific enrichment.

We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (Table 1). A number of metastatic enrichments were found in multiple clusters represented by pathways that were already enriched in a neighboring cluster. For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see supplemental figure S5) but not in 9 and 10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, Table 1). This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (Figure 2) and also in metastatic tumors of the sister groups 9 and 10, plus cluster 7. Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.

@@

### Correlation of microenvironment with tumor class

## Materials & Methods

## Discussion

### Classification of tumors independent of tissue-of-origin

@@

### Risk genes and tumor class

@@

### (Include discussions of other germline LOH)

@@

### Two-step and 3-strike

## References:

>>>>

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: "Peter T. Nguyen, Simon G. Coetzee, Daniel Lakeland and Dennis J. Hazelett"
output: 
  html_document:
    fig_captions: TRUE
    number_sections: FALSE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    code_folding: hide
    theme: flatly
bibliography: references.bib
---

@@

# Abstract
Cancer is a disease that affects metabolism and other basic cellular processes common to all cells. In spite of this, molecular signatures remain largely indistinguishable from tissue of origin, and general, unifying theories elude the field. To address this challenge we reduced omics mutation and expression profiling data to pathway memberships. We found that tumors can be classified according to pathway disruption in twelve distinct classes orthogonal to a tissue of origin (including histological subtypes), number of mutations and clinical stage. Moreover, known germline risk mutations such as BRCA1/2 are equally distributed among the classes. One class is distinguished by mutations in extracellular receptors and their ligands, particularly RTKs including the ERBB2/4, EGFR, Insulin Receptor and IGFR. This class is neatly subdivided into PI3K/Akt and Ras branches. Of these subclasses, the PI3K/Akt subgroup also has multiple lesions in immune-related pathways, including both innate immunity and T-cell receptor signaling. Another class is typified by mutations in the cellular response to hypoxia and defects in beta-catenin and hedgehog signaling. Intriguingly, we identify a novel class of mutations in proteasome component proteins that may be sufficient to cause compound defects across multiple signaling pathways in this class. A third class is typified by defects in metabolism, especially of RNA and protein biosynthesis. In addition to pathways that distinguish between the 3 major classes of tumor, we identify common mutations across tumors and stages, as well as pathways/mutations that distinguish between metastatic tumors and non-metastatic tumors. With these observations, we present a coherent picture of global tumor biology, in which genes and pathways can generally be assigned to one of three stages: initiation, expansion, and metastasis based on what groups the mutations are found. We discuss the implications for tumor biology and treatment.

# Introduction

@@

# Results

## Classification of tumors by pathway

@@

We noted intriguing similarities across different clusters, which motivated us to use dimensional reduction with UMAP [@McInnes2018]. Varying parameters and random seeds resulted in consistent, robust representations of the tumor field with 5 major groupings (Figure 1A). We call these tentatively by roman numerals I-V. While many of the clusters from the prior analysis remained close in proximity (note the colors in Figure \@ref(fig:umap-global-trends)A), some redistribution occurred, potentially reflecting important differences that could not be captured using the hierarchical method. Two of the five groupings were predominated by clusters from the hierarchical method distinguished chiefly in that one (cluster 5) had very few mutations and the other cluster was mutant for nearly every pathway (clusters 11 and 14). To determine whether the overall structure of the graph was driven by mutation counts, we examined the distribution of mutations (for pathway genes only). While there is a subtle trend, mostly visible for groups I and V (violin plot, Figure 1B), mutation count does not appear to be the dominant feature. To test whether the 5 groups represent progression by stage (which are determined clinically by size, though multiple inconsistent staging paradigms exist), we also projected these annotations onto the UMAP and found tumors of every stage, including metastatic tumors, in every grouping about equally (Figure \@ref(fig:umap-global-trends)C).

```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE}
set.seed(21)
plot(x=rnorm(1000), y=rnorm(1000))
```

The 15 hierarchical clusters were largely free of any tissue-of-origin bias, so we suspected that our 5 groups would be similarly independent from tissue of origin. To visualize tissue of origin relative to the groups we projected the annotations for 4 of the most common of the 23 cancer types in GDC, namely breast, prostate, ovarian and pancreatic cancers onto the UMAP, and did not observe any bias to group by these tissues of origin (Figures \@ref(fig:umap-global-trends)D). Full projections for all 23 cancers are presented in Supplementary Figure S2. Some cancers are defined by multiple subtypes of molecular or histologically distinguishing features, which are often the basis for different treatment therapies. One hypothesis to explain the lack of bias by cancer type is that the groupings are representative of these molecular subtypes. To probe this hypothesis, we projected annotations for each of the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive, normal-like, and lumenal A and B subtypes onto the UMAP. Perhaps surprisingly we did not detect any group biases for these or any other subtype annotations in GDC (see Supplementary Figure S3 for the full set of projections), suggesting that these subtypes are reflective of differences either in specialized cell-type-of-origin or other cancer risk mechanisms.

There are several well-known familial cancer-causing mutations that have been interrogated extensively for differences in basic biology, survival and treatment outcomes. However, the functions of these genes are related to risk factors such as genome stability, proof-reading and DNA damage repair, and telomere length. BRCA1/2 genes for example are key for DNA double-stranded break repair (ref) and confer risk for breast, prostate and ovarian cancers (refs). Since the mechanism of risk is thought to involve loss of heterozygosity, we projected the somatic mutations for BRCA1 and 2 genes onto the UMAP, but did not observe any segregation of these mutations by group (Figure \@ref(fig:umap-global-trends)E). We made similar projections for the MMR genes MSH2, MSH6, MLH, PMS1 and PMS2; BRIP1, RAD15, CHEK2 and APC. None of these genes exhibited any remarkable specificity with respect to group (except maybe APC??) (Supplementary Figure S4). To look at other risk factors such as maintenance of DNA methylation levels and telomere length we projected somatic mutations of the TET2 and TET3 genes, plus telomerase 1, 2, 3, and observed similar lack of segregation by group (Supplementary Figure S4).

## Determinants of Class

In order to discover whether any mutations of individual pathways corresponded to the major groupings (which we will hereafter refer to as classes) evident in the UMAP projection, we plotted the tumor samples harboring mutations for each of the 381 pathways used to produce it (Supplementary Figure S5), and then calculated enrichment for each pathway in-class relative to all other classes combined and ranked pathways for each class from highest to lowest enrichment. We visualized the enrichment as a heatmap, showing only the top significant pathways by x-fold enrichment (Figure \@ref(fig:group-pathway-enrichment)A). Using this approach, we identified 6? pathways associated with class II, 34 pathways primarily enriched in group III tumors, and 10 pathways enriched in class IV tumors. Neither group I nor class V had any significantly enriched pathways, perhaps low and high mutation rates in these groups, respectively. Group II was distinguished by metabolic pathways including RNA and protein biosynthesis (Figure \@ref(fig:group-pathway-enrichment)B). Most class II enriched pathways were not unique to this class but shared in one or more of the other classes, suggesting that this grouping reflects either greater heterogeneity or possibly a lower stage of tumor evolution. Class III typically has mutations in extracellular signaling, intracellular signaling, and immune-related pathways (Figure \@ref(fig:group-pathway-enrichment)C). Class IV had the highest pathway enrichment levels of the three, having mutations in hedgehog signaling, beta-catenin degradation, cellular response to hypoxia, regulation of cell cycle and apoptosis among others. As groups III and IV were the most distinct of the 5, we chose to investigate these two groups further.

```{r group-pathway-enrichment, fig.cap="Placeholder for figure 2", results=FALSE}
set.seed(28)
plot(density(rnorm(100)))
```

Group III contained interesting sub-patterns of enrichment, which we subdivided further into class IIIa and class IIIb sequentially from right following the global naming system. Common to classes IIIa and b were 20 or so pathways representing chiefly extracellular signaling receptors, including several cancer-associated receptor tyrosine kinases (RTK) such as EGFR, IGF1R, Insulin receptor, and ERBB2 and 4. In addition to these we observed SCF-Kit, DAP12, and Tie 2 signaling pathways, altogether suggesting that Class III tumors share mutations in extracellular signaling pathways generally. Since pathways like these may share many common components, we investigated whether tumors in this class typically have multiple mutations across signaling pathways, or whether a small number of critical shared components was driving the enrichment. We plotted mutations at the gene level, grouped in a supervised manner by pathway membership, and then performed unsupervised clustering of the tumors (Figure \@ref(fig:redundant-genes) or supplement figure?). We find… (please do this analysis).

```{r redundant-genes, fig.cap="Placeholder for figure 3", results=FALSE}
set.seed(29)
hist(rnorm(100))
```

@@

## Shared mutated genes/pathways

## Oncoprint stuff

## Metastatic _vs._ non-metastatic tumors

Following the same logic we used to investigate cluster-specific enrichment of pathways (Figure \@ref(fig:group-pathway-enrichment)) we applied this method to metastatic vs non-metastatic tumors, as we find that metastatic tumors are distributed across all 12 clusters (Figure \@ref(fig:umap-global-trends)C). Using all non-metastatic tumors as our background sample, we found very low levels of enrichment, less than 10%, in a handful of pathways (not shown). We reasoned that the clusters might be different enough to dilute cluster-specific enrichment.

We, therefore, calculated cluster-specific enrichment in metastatic tumors and found 99 enrichments across all clusters (Table 1). A number of metastatic enrichments were found in multiple clusters represented by pathways that were already enriched in a neighboring cluster. For example, “Beta-catenin independent WNT signaling” is enriched in non-metastatic samples of clusters 6, 7, 11 and 12 (see supplemental figure S5) but not in 9 and 10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the union of 9+10 (p < 10-3, Table 1). This is also true of “Erythropoietin activates RAS”, which is enriched in clusters 11 and 12 (Figure \@ref(fig:group-pathway-enrichment)) and also in metastatic tumors of the sister groups 9 and 10, plus cluster 7. Cluster 1, one of two clusters for which we could not identify any distinguishing pathways (cluster 8 being the other), had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 & 7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic samples were enriched for two key neutrophil pathways, “Neutrophil Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to metastases tend toward complementary mutations that are otherwise specific to non-metastatic tumors of other clusters.

@@

## Correlation of microenvironment with tumor class

# Materials & Methods

# Discussion

## Classification of tumors independent of tissue-of-origin

@@

## Risk genes and tumor class

@@

## (Include discussions of other germline LOH)

@@

## Two-step and 3-strike

# References
